---
title: "Analysis of GSE137001"
author: "James Dao"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    df_print: paged
    theme: cerulean
bibliography: refs/references.bib
csl: refs/addiction-biology.csl
link-citations: yes
---

```{r}
Sys.setenv(`_R_S3_METHOD_REGISTRATION_NOTE_OVERWRITES_` = "false")

suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
})

# Load relevant data
source(here("3_analysis", "data", "get_lfc_dfs.R"))
load(here("3_analysis", "data", "lfc_dfs.RData"))
source(here("3_analysis", "data", "get_enrichr_over.R"))
load(here("3_analysis", "data", "enrichr_over.RData"))

# knit options
options(knitr.duplicate.label = "allow")
```

# Intro

We will analyse bulk RNA-Seq data ([GSE137001](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE137001)) referenced by Velychko et al. (2019)[@Velychko2019], following these rough steps:

- Reprocess the raw data ([PRJNA564252](https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA564252&o=acc_s%3Aa)) using sra-tools, salmon, Snakemake and tximport
  - See the `2_reprocess` directory for details on the workflow
- Using MEFs as the base reference, perform DEG analysis with DESeq2 on every other timepoint
- Select significant over-expressed genes and pass them to the Enrichr browser tool
- Plot and tabulate DEGs (volcano plot and data table)


# Pathway Heatmaps (Enrichr Scores) {.tabset}

For each timepoint pair, we sent significant DEGs to [Enrichr](https://maayanlab.cloud/Enrichr/). A DEG is significant if it's adjusted p-value is less than 0.01, and absolute log2 fold chance is greater than 0.5.

```{r}
# Extract metrics from enrichr results
# GO:0036500 is commented out as it didn't show up in the enrichr results for all timepoints
# pathways <- c("GO:0009267", "GO:0034198", "GO:0036003", "GO:0036500", "GO:1990928", "GO:0034599", "GO:0034976", "GO:0016236", "GO:0010506")
pathways <- c("GO:0009267", "GO:0034198", "GO:0036003", "GO:1990928", "GO:0034599", "GO:0034976", "GO:0016236", "GO:0010506")

odds_ratios <- lapply(enrichr_over, function(res) {
  or <- res[["GO_Biological_Process_2021"]] %>%
    filter(str_detect(Term, paste(pathways, collapse="|"))) %>%
    arrange(Term) %>%
    pull(Odds.Ratio)
  
  return(or)
})
odds_ratios <- as.data.frame(odds_ratios)

neglog10_pvals <- lapply(enrichr_over, function(res) {
  pval <- res[["GO_Biological_Process_2021"]] %>%
    filter(str_detect(Term, paste(pathways, collapse="|"))) %>%
    arrange(Term) %>%
    mutate(neglog10 = -log10(Adjusted.P.value)) %>% 
    pull(neglog10)
  
  return(pval)
})
neglog10_pvals <- as.data.frame(neglog10_pvals)

combined_scores <- lapply(enrichr_over, function(res) {
  c_score <- res[["GO_Biological_Process_2021"]] %>%
    filter(str_detect(Term, paste(pathways, collapse="|"))) %>%
    arrange(Term) %>%
    pull(Combined.Score)
  
  return(c_score)
})
combined_scores <- as.data.frame(combined_scores)

row.names(odds_ratios) <- enrichr_over[[1]][["GO_Biological_Process_2021"]] %>%
  filter(str_detect(Term, paste(pathways, collapse="|"))) %>%
  arrange(Term) %>%
  mutate(Term = str_wrap(Term, width = 60)) %>% 
  pull(Term)

row.names(neglog10_pvals) <- row.names(odds_ratios)
row.names(combined_scores) <- row.names(odds_ratios)
```

## Odds Ratio

```{r}
pheatmap::pheatmap(
  odds_ratios,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = "row",
)
```

## -log10(adjusted p-value)

```{r}
pheatmap::pheatmap(
  neglog10_pvals,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = "row",
)
```

## Combined Score

```{r}
pheatmap::pheatmap(
  combined_scores,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = "row",
)
```


# Volcano Plots and DEG Tables {.tabset}

```{r, echo=FALSE, results='asis', cache=FALSE}
res <- lapply(names(lfc_dfs), function(x) {
  knitr::knit_child(
    here("3_analysis", "template.Rmd"),
    envir = environment(),
    quiet = TRUE
  )
})
cat(unlist(res), sep = '\n')
```

---

# References

<div id="refs"></div>


