---
title: "Cell Compass EDA"
author: "James Dao"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
    theme: cerulean
---

Links for convenience:

- DESeq2 Vignette
  - https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
- Original Paper
  - Short: https://pubmed.ncbi.nlm.nih.gov/32459171/
  - Full: https://www.ncbi.nlm.nih.gov/pmc/articles/pmid/32459171/


```{r include=FALSE}
library(here)
library(EnsDb.Hsapiens.v86)
library(DESeq2)
library(pheatmap)
library(tidyverse)
```


```{r message=FALSE}
# Load raw count data
raw_counts_path <- here("data", "counts.Rdata")

if (!file.exists(raw_counts_path)) {
  source(here("data", "get_data.R"))
}
load(raw_counts_path)
```


```{r}
# Add cell grouping used by paper (mentioned in the description of fig. 1D)
metadata <- metadata %>%
  mutate(cell_group = case_when(
    cell_type == "Fetal fibroblast" ~ "Fibroblast",
    cell_type == "Adult - Primary skin" ~ "Fibroblast",
    cell_type == "Embryonic stem cell" ~ "HUES_iPSCs",
    cell_type == "Induced pluripotent stem cell" ~ "HUES_iPSCs",
    cell_type == "Neural progenitor cell" ~ "NPCs",
    cell_type == "neuron" ~ "Neurons"
  )) %>%
  mutate(deg_group = case_when(
    source_name == "Skin" ~ "adult_primary_skin",
    source_name == "iPSC-derived neurons" ~ "iPSC_derived_neurons",
    TRUE ~ "untested"
  ))


# %>% mutate(  # So packages won't complain about spaces
#     source_name = gsub("[ -]", "_", source_name)
#   )

# Get gene symbols to convert from Ensembl IDs
ens2sym <- AnnotationDbi::select(
  EnsDb.Hsapiens.v86,
  keys = keys(EnsDb.Hsapiens.v86),
  columns = c("SYMBOL")
)
```


```{r message=FALSE, warning=FALSE}
# Create DESeq data objects for further analysis
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ deg_group  # not 100% sure about this
)
dds <- DESeq(dds, quiet = TRUE)

vsd <- vst(dds, blind = TRUE)
```

# Recreating Figure 1D

- We have decent distinct expression patterns and within-group uniformity
- However, compared to the original figure, our `HUES_iPSCs` and `Neurons` clusters seemed to be swapped
  - Possibly caused by using different underlying PCA algo
  - Plausibly the same plot, just shown at a different orientation about the 3rd (or higher) dimensions

```{r}
pca_data <- plotPCA(
  vsd,
  intgroup = c("cell_group"),
  returnData = TRUE,
  ntop = 1000  # Paper says "top 1000 miRNAs with the highest variance across all samples"
)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = cell_group)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()
```

# Recreating Figure 1G

Pretty much matches the paper.

```{r}
# Use the same ordering aa fig 1G in the original paper
colname_order <- c(
  "GSM2705984", "GSM2705985", "GSM2705983", "GSM2705987", "GSM2705988",
  "GSM2705989", "GSM2705990", "GSM2705986", "GSM2705992", "GSM2705993",
  "GSM2705994", "GSM2705995", "GSM2705991", "GSM2705997", "GSM2705998",
  "GSM2705999", "GSM2706000", "GSM2705996"
)

# Use selected genes from paper
neur_genes <- c(
  "NCAM1", "MAP2", "PAX6", "NEUROG2", "NEUROG1",
  "DCX", "TUBB3", "SOX2", "POU5F1", "NANOG"
)

rowname_order <- counts[matrixStats::rowSums2(counts) != 0, ] %>%
  as.data.frame() %>%
  rownames_to_column("GENEID") %>%
  left_join(ens2sym, by = "GENEID") %>%
  filter(SYMBOL %in% neur_genes) %>%
  arrange(match(SYMBOL, neur_genes)) %>%
  pull(GENEID)

pheatmap(
  assay(vsd)[rowname_order, colname_order],
  labels_row = neur_genes,
  annotation_col = select(metadata, cell_group),
  fontsize_col = 8,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

# Enrichr

## Compute log fold changes

Expression levels (measured by log2FoldChange) are computed using `iPSC_derived_neurons_vs_adult_primary_skin`.

- `iPSC_derived_neurons` uses 4 out of the 5 given neuron samples (HUES-derived neuron sample is excluded)
- `adult_primary_skin` uses 2 out of the 3 given fibroblast samples (fetal fibroblast sample is excluded)

Metadata link for convenience: https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA394766&o=acc_s%3Aa
  
```{r message=FALSE, warning=FALSE}
# Perform LFC shrinkage and join gene symbols
res_lfc <- lfcShrink(dds,
    coef = "deg_group_iPSC_derived_neurons_vs_adult_primary_skin",
    type = "apeglm", quiet = TRUE
  ) %>%
  as.data.frame() %>%
  rownames_to_column("GENEID") %>%
  left_join(ens2sym, by = "GENEID") %>%
  column_to_rownames("GENEID")
```

## Enrichr links v1

Over-expressed: https://maayanlab.cloud/Enrichr/enrich?dataset=e90d695972ab43ae34759f1b3a2d2e18

- My layman eyes see a decent amount of pathways related to brain stuff, and a bunch of other pathways I don't understand

Under-expressed: https://maayanlab.cloud/Enrichr/enrich?dataset=a7475202003c579570190f320f84a13a

- I see some pathways related to aging/stress:
  - Lysosome
  - Proteins Involved in Osteoarthritis
  - Proteins Involved in Myocardial Ischemia
  - reactive oxygen species degradation Homo sapiens DETOX1-PWY-1

```{r}
# Get overexpressed symbols for Enrichr
res_lfc %>%
  drop_na() %>%
  filter(padj < 0.01) %>%
  filter(log2FoldChange > 0) %>%
  select(SYMBOL) %>%
  write_csv("pos.txt", quote = "none", col_names = FALSE)

# Get underexpressed symbols for Enrichr
res_lfc %>%
  drop_na() %>%
  filter(padj < 0.01) %>%
  filter(log2FoldChange < 0) %>%
  select(SYMBOL) %>%
  write_csv("neg.txt", quote = "none", col_names = FALSE)
```






